Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-04 wpe.tex
minor edits to wpe, with a note at the end

The goal of the section is to describe our solution to the challenging
problem of estimating the entropy of an arbitrary, real-valued time
series.  This approach draws upon methods and results from a variety
of fields including time-series analysis, dynamical systems, and
stochastic processes.

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  Time-series consisting of IID 
random variables, such as white noise, have maximal entropy rates, 
whereas highly structured time-series, for example periodic, have very
low or zero entropy rate. A time series with a high entropy rate is 
almost completely unpredictable; conversely one with low entropy rate
is often quite predictable. This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy, and
the more difficult prediction becomes.  The KS entropy is defined as
the supremum of the Shannon entropy rates of all
partitions~\cite{petersen1989}. The partition that achieves this
supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present which is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
the more redundancy in a signal, the more predictable it is.  
Redundancy is critical to prediction, in general.  And the specific 
\emph{form} of the redundancy dictates whether a particular prediction 
method will work well or poorly on the corresponding signal.  A linear 
method cannot detect or make use of nonlinear redundancy, for 
instance.  This issue is central to the claims in this paper and the 
discussion in the following section.

Utilizing Shannon entropy rate as a measure of temporal complexity is
not a new idea \cite{Shannon1951, mantegna1994linguistic}, but
previous approaches required categorical data: $x_i \in \mathcal{S}$
for some finite or countably infinite \emph{alphabet} $\mathcal{S}$.
Data taken from real-world systems is, however,
effectively\footnote{Measurements from finite-precision sensors are
  discrete, but data from modern high-resolution sensors are, for the
  purposes of entropy calculations, effectively continuous.}
real-valued.  To analyze real-valued data using a method that requires
categorical values, one must discretize the data---typically by
binning.  Unfortunately, this is rarely a good solution, as the
binning of the values introduces spurious dynamics~\cite{bollt2001}.
The field of symbolic dynamics offers discretization methods that do
not disturb the intrinsic behavior.  These methods are, however,
fragile in the face of noise; worse yet, they require knowledge of the
underlying system.  This is not useful in our context, where we want
to measure the entropy to quantify predictability.

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, so long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system. This transitive chain of equalities, from permutation entropy 
to Shannon entropy rate via the KS entropy, then allows us to
approximate the redundancy --- being the dual of the Shannon entropy 
rate --- of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about 
the fixed points driving it to be $\approx 1$, in some sense hiding
the fact that the signal is actually quite structured. The weighted 
permutation entropy on the other hand would be dominated by the 
switching rather than by the stochastic fluctuations. To accomplish 
this, the \emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

The weighted probability of a permutation is defined as:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell) \cdot \delta(\phi(x_{i+1}^\ell), \pi) }{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell)}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series. Using the
standard form of an entropy, the weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, making $0
\le \textrm{WPE} \le 1$.  This normalization is used in all the
results that follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration here is that the value be large enough that
forbidden ordinals are discovered, yet small enough that reasonable
statistics over the ordinals are
gathered.  If an average of 100 counts per ordinal is considered to be
sufficient, for instance, then $\ell =
\operatornamewithlimits{argmax}_{\hat{\ell}} \{ N \gtrapprox 100 \hat{\ell}! \}$.  In the
literature, $3 \le \ell \le 6$ is a standard choice---generally
without any formal justification.  In theory, the permutation entropy
should reach an asymptote with increasing $\ell$, but that can require
an arbitrarily long time series. In practice, what one should do is
calculate the \emph{persistent} permutation entropy by increasing
$\ell$ until the result converges, but data length issues can intrude
before that convergence is reached.  We used this approach to choose
$\ell = 6$ for the experiments in the following section.  This value
represents a good balance between accurate ordinal statistics and
finite-data effects.

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities 
between the two research projects.  Haven {\sl et al.} utilize the 
relative entropy to quantify how much better a distribution over 
system states evolved from small ensembles to predict the future of a 
dynamical system compared to its asymptotic distribution. Our work 
quantifies the predictability of a single observed time series using 
weighted permutation entropy and makes no assumptions about the 
underlying dynamics. These approaches are very different: Haven's 
compares two distributions, the average behavior versus the behaviors 
that can be seen given a reasonably known state in the past; ours 
attempts to quantify how well an appropriate predictor will do for a 
given time-series without any understanding of how that time-series
was generated.

>>>>

Estimating the entropy of an arbitrary, real-valued time series is a
challenging problem.  Our approach to this problem, which is the topic
of this section, draws upon methods and results from a variety of
fields including time-series analysis, dynamical systems, and
stochastic processes.

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  Time-series consisting of i.i.d.
random variables, such as white noise, have maximal entropy rates, 
whereas highly structured time-series, for example periodic, have very
low or zero entropy rate. A time series with a high entropy rate is 
almost completely unpredictable; conversely, one with low entropy rate
is often quite predictable. This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy, and
the more difficult prediction becomes.  The KS entropy is defined as
the supremum of the Shannon entropy rates of all
partitions~\cite{petersen1989}. The partition that achieves this
supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present that is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
the more redundancy in a signal, the more predictable it is.  
Redundancy is critical to prediction, in general.  And the specific 
\emph{form} of the redundancy dictates whether a particular prediction 
method will work well or poorly on the corresponding signal.  A linear 
method cannot detect or make use of nonlinear redundancy, for 
instance.  This issue is central to the claims in this paper and the 
discussion in the following section.

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems is, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where we want to measure the entropy to
quantify predictability.

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, so long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system. This transitive chain of equalities, from permutation entropy
to Shannon entropy rate via the KS entropy, then allows us to
approximate the redundancy---being the dual of the Shannon entropy
rate---of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about
the fixed points driving it to $\approx 1$, which in some sense hides
the fact that the signal is actually quite structured.  In this
situation, the weighted permutation entropy measures on switching
rather than the stochastic fluctuations. To accomplish this, the
\emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

The weighted probability of a permutation is defined as:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell) \cdot \delta(\phi(x_{i+1}^\ell), \pi) }{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell)}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series. Using the
standard form of an entropy, the weighted permutation entropy is:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, making $0
\le \textrm{WPE} \le 1$.  This normalization is used in all the
results that follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration in that choice is that the value be large enough
that forbidden ordinals are discovered, yet small enough that
reasonable statistics over the ordinals are gathered.  If an average
of 100 counts per ordinal is considered to be sufficient, for
instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
6$ is a standard choice---generally without any formal justification.
In theory, the permutation entropy should reach an asymptote with
increasing $\ell$, but that can require an arbitrarily long time
series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $\ell$ until the
result converges, but data length issues can intrude before that
convergence is reached.  We used this approach to choose $\ell = 6$
for the experiments in the following section.  This value represents a
good balance between accurate ordinal statistics and finite-data
effects.

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects.  Haven {\sl et al.} utilize the
relative entropy to quantify how much better a distribution over
system states evolved from small ensembles to predict the future of a
dynamical system compared to its asymptotic distribution. Our work
quantifies the predictability of a single observed time series using
weighted permutation entropy and makes no assumptions about the
underlying dynamics. {\color{red} These approaches are very different:
  Haven's compares two distributions, the average behavior versus the
  behaviors that can be seen given a reasonably known state in the
  past; ours attempts to quantify how well an appropriate predictor
  will do for a given time-series without any understanding of how
  that time-series was generated. [[Seems to me like the first part of
      this paragraph made the point.  Is this second set of stuff
      needed?]]}
